---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(tibble.print_min = 5, tibble.print_max = 5)
```
# mpraprofiler

<!-- badges: start -->
<!-- badges: end -->

The goal of mpraprofiler is to create a simple tool to perform allelic differential expression analysis for **massively parallel reporter assays** data

## Installation

You can install the the development version version of mpraprofiler from gitlab with. It is still too early for this package to be public. You can use `getPass` package to pass the authorization for gitLab:

``` {r, eval = FALSE}
# install.packages("devtools")
# install.packages("getPass")
devtools::install_git("https://tfwebdev.research.cchmc.org/gitlab/lux2ht/mpraprofiler.git", ,credentials = git2r::cred_user_pass("lux2ht", getPass::getPass()))
```

## Example
please refer to **vignettes** for details
